---
title: "1_how_to_make_a_heatmap"
author: "Ming Tang"
date: "5/2/2023"
output: html_document
editor_options: 
  chunk_output_type: console
---


Making heatmap is an essential skill for any computational biologist.


### load the libraries 

```{r}
library(ComplexHeatmap)

```

### make dummy data 

```{r}

h1 <- c(10,20,10,20,10,20,10,20)
h2 <- c(20,10,20,10,20,10,20,10)

l1 <- c(1,3,1,3,1,3,1,3)
l2 <- c(3,1,3,1,3,1,3,1)

mat <- rbind(h1,h2,l1,l2)
colnames(mat)<- paste0("timepoint_", 1:8)
mat
```


visualize the data 

```{r}
par(mfrow =c(1,1), mar=c(4,4,1,1))
plot(1:8,rep(0,8), ylim=c(0,35), pch="", xlab="Time", ylab="Gene Expression")

for (i in 1:nrow(mat)) {
lines(1:8,mat[i,], lwd=3, col=i)
}

legend(1,35,rownames(mat), 1:4, cex=0.7)
```

### Making a heatmap is easy!

```{r}
Heatmap(mat, cluster_columns = FALSE)

quantile(mat, c(0, 0.1,0.5, 0.9))
```


change color mapping 

```{r}
col_fun<- circlize::colorRamp2(c(0, 3, 20), c("blue", "white", "red"))

Heatmap(mat, cluster_columns = FALSE, col = col_fun)
```


outlier 

```{r}
mat2<- mat
mat2[1,1]<- 1000

Heatmap(mat2, cluster_columns = FALSE)

Heatmap(mat2, cluster_columns = FALSE, col = col_fun)

```

Let's scale the data/gene expression level across columns (time points) first

```{r}
scaled_mat<- t(scale(t(mat)))

scaled_mat
?Heatmap
Heatmap(scaled_mat, cluster_columns = FALSE)
```

Note, after scaling, h1 and l1 are close to each other!

### understand clustering?

define distances 

```{r}
?dist

d<- dist(mat)

d
```


```{r}
?hclust

hclust(d)


plot(hclust(d))
```


After scaling 
```{r}

d2<- dist(scaled_mat)

plot(hclust(d2))
```

Key takeaways:
1. color mapping is critical
2. scaling is critical 
3. making heatmap is easy, but better to understand the details.







