---
title: "differential_expression_heatmap"
author: "Ming Tang"
date: "5/9/2023"
output: html_document
editor_options: 
  chunk_output_type: console
---

Let's use a real example https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE197576

How to download the files from ftp https://www.ncbi.nlm.nih.gov/geo/info/download.html

https://ftp.ncbi.nlm.nih.gov/geo/series/GSE197nnn/GSE197576/suppl/

Alternative use GEOquery https://bioconductor.org/packages/release/bioc/html/GEOquery.html

```{bash eval=FALSE}
wget https://ftp.ncbi.nlm.nih.gov/geo/series/GSE197nnn/GSE197576/suppl/GSE197576_raw_gene_counts_matrix.tsv.gz

csvtk pretty -t  GSE197576_raw_gene_counts_matrix.tsv.gz | less -S

csvtk headers -t   GSE197576_raw_gene_counts_matrix.tsv.gz

csvtk cut -t -f1,2,3,12,13 GSE197576_raw_gene_counts_matrix.tsv.gz| head

csvtk cut -t -f1,2,3,12,13 GSE197576_raw_gene_counts_matrix.tsv.gz > raw_counts.tsv
```

Get csvtk at https://github.com/shenwei356/csvtk

### read the data into R and make a DESeq2 object 

follow the tutorial http://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html

```{r}
library(dplyr)
library(readr)
library(here)
library(DESeq2)

raw_counts<- read_tsv(here("data/raw_counts.tsv"))

raw_counts_mat<- raw_counts[, -1] %>% as.matrix

head(raw_counts_mat)

rownames(raw_counts_mat)<- raw_counts$gene
head(raw_counts_mat)

```

Make a sample sheet 

```{r}
coldata<- data.frame(condition = c("normoxia", "normoxia", "hypoxia", "hypoxia"))

rownames(coldata)<- colnames(raw_counts_mat)

coldata
```

Make a DEseq2 object

```{r}
all(rownames(coldata) == colnames(raw_counts_mat))

dds <- DESeqDataSetFromMatrix(countData = raw_counts_mat,
                              colData = coldata,
                              design = ~ condition)
dds <- DESeq(dds)
res <- results(dds, contrast = c("condition", "hypoxia", "normoxia"))

res %>%
  as.data.frame() %>%
  arrange((padj), desc(log2FoldChange)) %>%
  head(n=30)


significant_genes<- res %>%
  as.data.frame() %>%
  filter(padj <=0.01, abs(log2FoldChange) >= 2) %>% 
  rownames()


significant_genes
```


### PCA analysis

```{r}
vsd <- vst(dds, blind=FALSE)

plotPCA(vsd, intgroup=c("condition"))
```


Plot PCA by ourselves.

```{r}
vsd <- vst(dds, blind=FALSE)
head(assay(vsd), 3)

normalized_counts<- assay(vsd) %>% 
  as.matrix()

pca_prcomp<- prcomp(t(normalized_counts), center = TRUE, scale. = FALSE)

names(pca_prcomp)
pca_prcomp$x

PC1_and_PC2<- data.frame(PC1=pca_prcomp$x[,1], PC2= pca_prcomp$x[,2], 
                         type = rownames(pca_prcomp$x))

## plot PCA plot
library(ggplot2)

ggplot(PC1_and_PC2, aes(x=PC1, y=PC2, col=type)) + 
  geom_point() + 
  geom_text(aes(label = type), hjust=0, vjust=0) +
  coord_fixed()
```

It is not exactly the same, what's going on? 


```{r}
?plotPCA #using the top 500 most variable genes

```

https://github.com/mikelove/DESeq2/blob/48b80aaac5efd4b9e0d054fc1e4a6e1fa78e782a/R/plots.R#LL245C71-L245C71


### Make a perfect heatmap

```{r}
library(ComplexHeatmap)

significant_mat<- normalized_counts[significant_genes, ] 

Heatmap(t(scale(t(significant_mat))))

```


Yeah, you get this perfect looking heatmap because we select the genes that are different. So, no surprise at all!


```{r}
coldata

col_anno <- HeatmapAnnotation(df = coldata, 
                             col = list( condition = c("hypoxia" = "red", "normoxia" = "blue")))


Heatmap(t(scale(t(significant_mat))), 
        top_annotation = col_anno,
        show_row_names = FALSE,
        name = "scaled normalized\nexpression")
```

why scaling is important?


```{r}
Heatmap(significant_mat, 
        top_annotation = col_anno,
        show_row_names = FALSE,
        name = "scaled normalized\nexpression")
```

