---
title: "volcano plot with ggrepel"
author: "Ming Tang"
date: "5/16/2023"
output: html_document
editor_options: 
  chunk_output_type: console
---


Continue with https://github.com/crazyhottommy/compbio_tutorials/blob/main/scripts/02_differential_expression_heatmap.Rmd

### read the data into R and make a DESeq2 object 

follow the tutorial http://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html

```{r}
library(dplyr)
library(readr)
library(here)
library(DESeq2)
library(ggplot2)

raw_counts<- read_tsv(here("data/raw_counts.tsv"))

raw_counts_mat<- raw_counts[, -1] %>% as.matrix

head(raw_counts_mat)

rownames(raw_counts_mat)<- raw_counts$gene
head(raw_counts_mat)

```

Make a sample sheet 

```{r}
coldata<- data.frame(condition = c("normoxia", "normoxia", "hypoxia", "hypoxia"))

rownames(coldata)<- colnames(raw_counts_mat)

coldata
```

Make a DEseq2 object

```{r}
all(rownames(coldata) == colnames(raw_counts_mat))

dds <- DESeqDataSetFromMatrix(countData = raw_counts_mat,
                              colData = coldata,
                              design = ~ condition)
dds <- DESeq(dds)
res <- results(dds, contrast = c("condition", "hypoxia", "normoxia"))

head(res)
```

### Make a volcano plot

what is a volcano plot?

It is a scatter plot:
x-axis is the log2Fold change

y-axis is -log10(p-value)


```{r}
res %>% 
  as.data.frame() %>%
  ggplot(aes(x = log2FoldChange, y = -log10(pvalue))) +
  geom_point()
  
```

hmm, what are the points on the top?

```{r}
res %>%
  as.data.frame() %>%
  arrange((padj), desc(log2FoldChange)) %>%
  head(n = 30)
```

A basic volcano plot

```{r}
res %>% 
  as.data.frame() %>%
  ggplot(aes(x = log2FoldChange, y = -log10(pvalue))) +
  geom_point() +
  theme_bw(base_size = 14)
```

How to label the genes?

```{r}

res %>% 
  as.data.frame() %>%
  tibble::rownames_to_column(var = "gene") %>%
  filter(!stringr::str_detect(gene, "LOC")) %>% 
  filter(abs(log2FoldChange)>=4, padj <= 0.001) %>%
  dim()


res_sig<- res %>% 
  as.data.frame() %>%
  tibble::rownames_to_column(var = "gene") %>%
  filter(!stringr::str_detect(gene, "LOC")) %>% 
  filter(abs(log2FoldChange)>=4, padj <= 0.001) 
  

res %>% 
  as.data.frame() %>%
  ggplot(aes(x = log2FoldChange, y = -log10(pvalue))) +
  geom_point() +
  geom_label(data = res_sig, aes(label = gene))+
  theme_bw(base_size = 14)
```


the labels are overlapping, let's improve by ggrepel
```{r}
#install.packages("ggrepel")
library(ggrepel )


res %>% 
  as.data.frame() %>%
  ggplot(aes(x = log2FoldChange, y = -log10(pvalue))) +
  geom_point() +
  ggrepel::geom_label_repel(data = res_sig, aes(label = gene))+
  theme_bw(base_size = 14)
```


Let's color the points 

```{r}
res2<- res %>% 
  as.data.frame() %>%
  tibble::rownames_to_column(var = "gene") %>%
  mutate(sig = case_when(
    !stringr::str_detect(gene, "LOC") & abs(log2FoldChange)>=4 & 
      padj <= 0.001 ~ "sig",
    TRUE ~ "not sig"
  ))
  
head(res2)

ggplot(res2, aes(x = log2FoldChange, y = -log10(pvalue))) +
  geom_point(aes(color = sig)) +
  ggrepel::geom_label_repel(data = res_sig, aes(label = gene))+
  theme_bw(base_size = 14)
```


fix the color of the points 

```{r}
ggplot(res2, aes(x = log2FoldChange, y = -log10(pvalue))) +
  geom_point(aes(color = sig)) +
  scale_color_manual(values = c("blue", "red")) +
  ggrepel::geom_label_repel(data = res_sig, aes(label = gene))+
  theme_bw(base_size = 14)
```

add horizontal and vertical lines
```{r}
ggplot(res2, aes(x = log2FoldChange, y = -log10(pvalue))) +
  geom_point(aes(color = sig)) +
  scale_color_manual(values = c("blue", "red")) +
  ggrepel::geom_label_repel(data = res_sig, aes(label = gene))+
  geom_hline(yintercept = 100, linetype = 2, color = "red") +
  geom_vline(xintercept = c(-4, 4), linetype = 2, color = "red")+
  theme_bw(base_size = 14)
```

Enhanced volcanoplot: https://bioconductor.org/packages/devel/bioc/vignettes/EnhancedVolcano/inst/doc/EnhancedVolcano.html

