---
title: "05_find_tissue_specific_genes_human_protein_atlas"
author: "Ming Tang"
date: "6/15/2023"
output: html_document
editor_options: 
  chunk_output_type: console
---

http://www.bioconductor.org/packages/devel/bioc/vignettes/HPAanalyze/inst/doc/b_HPAanalyze_indepth.html


```{r}
#BiocManager::install("HPAanalyze")
library(tidyverse)
library(HPAanalyze)
```

Most of the time, you will only need the “histology” datasets, which contain normal_tissue, pathology (basically cancers) and subcellular_location.

```{r}
downloadedData <- hpaDownload(downloadList='histology')
```

The `normal_tissue` dataset contains information about protein expression profiles in human tissues based on IHC staining. The datasets contain six columns: ensembl (Ensembl gene identifier); gene (HGNC symbol), tissue (tissue name); cell_type (annotated cell type); level (expression value); reliability (the gene reliability of the expression value)

```{r}

names(downloadedData)

downloadedData$normal_tissue %>%
  head()

table(downloadedData$normal_tissue$level)

downloadedData$normal_tissue %>%
  filter(cell_type == "smooth muscle cells") %>%
  janitor::tabyl(tissue, cell_type)

downloadedData$normal_tissue %>%
  pivot_wider(names_from = c("tissue", "cell_type"), values_from = "level") %>%
  View()

data<- downloadedData$normal_tissue %>%
  pivot_wider(names_from = c("tissue", "cell_type"), values_from = "level") %>%
  filter(`smooth muscle_smooth muscle cells` %in% c("High", "Medium"))

View(data)
```

what are the membrane genes

```{r}
head(downloadedData$subcellular_location)

table(downloadedData$subcellular_location$main_location) %>% 
  sort()

memberane_genes<- downloadedData$subcellular_location %>% 
  filter(str_detect(main_location, "Plasma membrane")) 

inner_join(data, memberane_genes, by = c("ensembl" = "ensembl")) 
```


Make a heatmap 

```{r}
library(ComplexHeatmap)

data2<- data %>% 
  filter(gene %in% memberane_genes$gene)
  
mat<- as.matrix(data2[, -c(1,2,3)])
rownames(mat)<- data2$gene

ComplexHeatmap::Heatmap(mat,
                        show_row_names = FALSE,
                        show_column_names = FALSE)
```



