---
title: "singleCellExperiment"
author: "Ming Tang"
date: '2024-06-01'
output: html_document
editor_options: 
  chunk_output_type: console
---

### introduction to singleCellExperiment 

references:

https://www.bioconductor.org/packages/release/bioc/vignettes/SingleCellExperiment/inst/doc/intro.html


```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("scater")

BiocManager::install("SingleCellExperiment")

library(SingleCellExperiment)
library(scater)
```

https://ftp.ncbi.nlm.nih.gov/geo/series/GSE126nnn/GSE126030/suppl/

```{bash eval=FALSE}
wget https://ftp.ncbi.nlm.nih.gov/geo/series/GSE126nnn/GSE126030/suppl/GSE126030_RAW.tar

tar -xvf GSE126030_RAW.tar
```


```{r}
library(tidyverse)
library(here)
counts_df<- read_tsv(here("data/GSM3589407_PP002swap.filtered.matrix.txt.gz"))

colnames(counts_df) %>% tail()

counts_df$...2466 %>% head()

table(counts_df$...2466, useNA="ifany")
```

remove the last column
```{r}
length(colnames(counts_df))
counts_df<- counts_df[, -2466]

counts_df[1:5, 1:5]

counts_mat<- counts_df[, -c(1,2)] %>%
  as.matrix()

rownames(counts_mat)<- counts_df$Gene

counts_mat[1:5, 1:5]
```

### Create a SingleCellExperiment Object 

```{r}
sce <- SingleCellExperiment(assays = list(counts = counts_mat))

sce
```
To access the count data we just supplied, we can do any one of the following:

```{r}
assay(sce, "counts")

assay(sce, "counts")[1:5, 1:5]

counts(sce)[1:5, 1:5]
```

One can access the slots with the `@` operator. This is considered bad practice as the class developers are free to alter the internal structure of the class, at which point any code using `@` may no longer work. Rather, it is best to use the provided **getter** functions like `assay()` and `counts()`

```{r}
sce@assays

sce@assays@data

sce@assays@data$counts[1:5, 1:5]
```

add more assays

```{r}
sce <- scuttle::logNormCounts(sce)
sce
```

```{r}
sce@assays@data$logcounts[1:5, 1:5]

assay(sce, "logcounts")[1:5, 1:5]

dim(logcounts(sce))
```

You may add the assay manually

```{r}
counts_1 <- counts(sce) + 1

assay(sce, "counts_1") <- counts_1 # assign a new entry to assays slot

assays(sce)

# not recommended way to add new assay data
sce@assays@data$counts_2<- counts(sce) + 2
assays(sce)
```

```{r}
assayNames(sce)
```

### add metadata

```{r}
sce <- scuttle::addPerCellQC(sce)


colData(sce)

coldata<- colData(sce)

identical(rownames(coldata), colnames(counts_mat))

sce <- scuttle::addPerFeatureQC(sce)
rowData(sce)
```

```{r}
rowRanges(sce)
```

### other metadata

the metadata slot, a named list of arbitrary objects. For example, say we have some favorite genes (e.g., highly variable genes) that we want to store inside of sce for use in our analysis at a later point. 

```{r}
my_genes <- c("gene_1", "gene_5")
metadata(sce) <- list(favorite_genes = my_genes)
metadata(sce)
```

### dimension reduction

```{r}
sce <- scater::logNormCounts(sce)

sce <- scater::runPCA(sce)

sce


reducedDim(sce, "PCA")[1:5, 1:5]

dim(reducedDim(sce, "PCA"))

sce <- scater::runTSNE(sce, perplexity = 0.1)

reducedDim(sce, "TSNE")[1:5, 1:2]

reducedDims(sce)


reducedDimNames(sce)
```

### add 

```{r}
u <- uwot::umap(t(logcounts(sce)), n_neighbors = 2)
reducedDim(sce, "UMAP_uwot") <- u

reducedDims(sce) # Now stored in the object.
```


```{r}
colLabels(sce) <- scran::clusterCells(sce, use.dimred="PCA")
table(colLabels(sce))
```

```{r}
scater::plotReducedDim(sce, dimred="TSNE")
```

