---
title: "09_intro_to_seurat_v5"
author: "Ming Tang"
date: '2024-06-06'
output: html_document
editor_options: 
  chunk_output_type: console
---

### Overview of Layers in Seurat V5

In Seurat V5, a Layer is a new abstraction that can encapsulate different modalities or features of the single-cell data. This allows users to seamlessly integrate and analyze data from various sources while retaining the ability to use familiar Seurat functions.

Key Concepts of Layers
Layered Data: Each layer can represent a modality (like RNA, protein, etc.) or a different version of data (like log-normalized versus raw counts).


```{r}
options(Seurat.object.assay.version = "v5")
library(Seurat)
library(dplyr)

# packageVersion("Seurat")

```

Load the example data
```{r}
# devtools::install_github('satijalab/seurat-data')
library(SeuratData)
InstallData("pbmc3k")
data("pbmc3k")
pbmc3k

pbmc3k<- UpdateSeuratObject(pbmc3k)

pbmc3k
```


```{r}
pbmc3k<- pbmc3k %>% 
  NormalizeData(normalization.method = "LogNormalize", scale.factor = 10000) %>%
  FindVariableFeatures(selection.method = "vst", nfeatures = 2000) %>%
  ScaleData() %>%
  RunPCA(verbose = FALSE) %>%
  RunUMAP(dims = 1:10, verbose = FALSE)
```

### Seurat V5

Seurat v5 assays store data in layers. These layers can store raw, un-normalized counts (layer='counts'), normalized data (layer='data'), or z-scored/variance-stabilized data (layer='scale.data').

```{r}
pbmc3k@meta.data %>% head()
pbmc3k@assays
pbmc3k@assays$RNA
pbmc3k[["RNA"]]

Layers(pbmc3k)

LayerData(pbmc3k, "counts")[1:5, 1:5]
pbmc3k@assays$RNA$counts[1:5, 1:5]
pbmc3k[["RNA"]]@counts[1:5, 1:5]


LayerData(pbmc3k, "data")[1:5, 1:5]


##use getter function
GetAssayData(object = pbmc3k, slot = 'data')[1:5, 1:5]

```



dimension reduction

```{r}
pbmc3k@reductions

pbmc3k@reductions$pca

pbmc3k@reductions$pca@cell.embeddings %>% head()

pbmc3k[["pca"]]
pbmc3k[["pca"]]@cell.embeddings %>% head()

## getter function

Embeddings(pbmc3k, "pca") %>% head()
Embeddings(pbmc3k, "umap") %>% head()

Loadings(pbmc3k, "pca") %>% head()

```

### convert between V4 and V5

```{r}
obj<- pbmc3k

#convert v5 assay to v4 assay within same object
obj[["RNA"]]<- as(obj[["RNA"]], Class= "Assay")

class(obj[["RNA"]])
# convert it back
obj[["RNA"]]<- as(obj[["RNA"]], Class= "Assay5")

class(obj[["RNA"]])

```

