---
title: "get gene length"
output: html_document
date: "2025-01-16"
editor_options: 
  chunk_output_type: console
---

```{r}
#install.packages("BiocManager")
#BiocManager::install("TxDb.Mmusculus.UCSC.mm9.knownGene")
#BiocManager::install("org.Mm.eg.db")

library(TxDb.Mmusculus.UCSC.mm9.knownGene)
library(org.Mm.eg.db)
```

Now, we can retrieve the gene data:
```{r}
txdb <- TxDb.Mmusculus.UCSC.mm9.knownGene
mm9_genes <- genes(txdb)
mm9_genes
```

To map the Entrez ID to gene symbols, we use: 

```{r}
gene_symbol <- AnnotationDbi::select(org.Mm.eg.db, keys = mm9_genes$gene_id, 
                                     columns = "SYMBOL", keytype = "ENTREZID")
```

Make sure the gene IDs match: 

```{r}
all.equal(mm9_genes$gene_id, gene_symbol$ENTREZID)

# Add gene symbols to the data: 
mm9_genes$symbol <- gene_symbol$SYMBOL
```

```{r}
width(mm9_genes)

df <- data.frame(EntrezID = mm9_genes$gene_id, 
                 Symbol = mm9_genes$symbol, 
                 Gene_length = width(mm9_genes))

head(df)
```

Why is this useful? When normalizing H3K36me3 signals (found in gene bodies), you can use gene lengths for proper normalization. Also, for RNA-seq, exon lengths are essential for calculating metrics like RPKM (though TPM is preferred now).

### exon lengths 

```{r}
exons<- exonsBy(txdb, by = "gene")
exons
```

This returns a GRangesList object and each element of the list is a GRanges containing all the exons for that gene.

Let’s calculate the total exon lengths for each gene by the width function

```{r}
width(exons)

head(sum(width(exons)))
```

```{r}
exon_len<- sum(width(exons)) %>%
      tibble::enframe(name = "ENTREZID", value = "exon_length")

head(exon_len)
```

